---
title: "Project 3 part 2"
author: "Isaac Johnson & Jesse DeBolt"
date: "04/11/2023"
output:
  html_document:
    df_print: paged
  pdf_document:
    df_print: kable
    fig_width: 11
    fig_height: 8
---


NOTE: We had some issues trying to merge our two documents, this document in additional to our original submit should be considered the project as a whole. Some EDA is included in part 1, the original submission as well as the PCA section and some modeling. This document contains the K-Means clustering section as well as some additional EDA and modeling.


## Setup
```{r message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(tidytext)
library(caret)
library(fastDummies)
library(randomForest)
library(broom)
# https://www.openml.org/d/1590

raw_income = read_csv("./openml_1590.csv", na=c("?"))

income = read_csv("./openml_1590.csv", na=c("?")) %>%
  drop_na() %>%
  mutate(income_above_50k = class==">50K") %>%
  select(-class) %>%
  dummy_cols(remove_selected_columns = T)

head(income)
```


##preprocessing before K-means clustering
```{r}
#preprocessing data
income_scaled <- income %>% 
    mutate(age=scale(age), fnlwgt=scale(fnlwgt), `education-num`=scale(`education-num`), `capital-gain`=scale(`capital-gain`), `capital-loss`=scale(`capital-loss`), `hours-per-week`=scale(`hours-per-week`))
```


##Basic K-means clustering
```{r}
#setting # of clusters to 3 (that's K)
kclust <- kmeans(income_scaled, centers = 3)
kclust$centers

```

```{r}
glance(kclust)
```


##Add Clusters to original dataset
```{r}
incomek <- augment(kclust,income_scaled)
head(incomek)
```

##Visualize Clusters
```{r}

incomek %>% 
  pivot_longer(c(age, fnlwgt, `education-num`),names_to = "feature") %>% 
  ggplot(aes(value, fill=.cluster))+
  geom_density(alpha=0.3)+
  facet_wrap(~feature)

incomek %>% 
  pivot_longer(c(`capital-gain`,`capital-loss`),names_to = "feature") %>% 
  ggplot(aes(value, fill=.cluster))+
  geom_density(alpha=0.3)+
  facet_wrap(~feature)

incomek %>% 
  pivot_longer(c(`hours-per-week`,workclass_Private),names_to = "feature") %>% 
  ggplot(aes(value, fill=.cluster))+
  geom_density(alpha=0.3)+
  facet_wrap(~feature)

incomek %>% 
  pivot_longer(c(`hours-per-week`,income_above_50k),names_to = "feature") %>% 
  ggplot(aes(value, fill=.cluster))+
  geom_density(alpha=0.3)+
  facet_wrap(~feature)

```

##Try different numbers of clusters
```{r}
kclusts <- tibble(k = 1:9) %>%
  mutate(
    kclust = map(k, ~kmeans(income_scaled, .x)),
    glanced = map(kclust, glance),
    augmented = map(kclust, augment, income_scaled)
  )


```

##Plot the different clusters on two axes
```{r}
assignments <- kclusts %>% 
  unnest(augmented)


ggplot(assignments, aes(`hours-per-week`, `education-num`)) +
  geom_point(aes(color = .cluster), alpha=0.3) + 
  facet_wrap(~ k)
```

##Look at improvement in within-cluster error
```{r}
#can still look for elbow (looks like about 7)
clusterings <- kclusts %>%
  unnest(glanced, .drop = TRUE)

ggplot(clusterings, aes(k, tot.withinss)) +
  geom_line()
```


```{r}
# Run k-means with the optimal number of clusters
optimal_k <- 7
kclust_optimal <- kmeans(income_scaled, centers = optimal_k)

# Check the cluster centroids
kclust_optimal$centers

```

```{r}
# Run k-means with 7 clusters
kclust <- kmeans(income_scaled, centers = 7)

# Print the cluster centers
print(kclust$centers)

```

##Outliers lof
```{r}
library(dbscan)
lof <- lof(income, minPts = 10)
summary(lof)

hist(lof, breaks = 10, main = "LOF (minPts = 10)")

plot(sort(lof), type = "l",  main = "LOF (minPts = 10)",
  xlab = "Points sorted by LOF", ylab = "LOF")

#capital gain/loss
plot(select(income, c("capital-gain", "capital-loss")), pch = ".", main = "LOF (minPts = 10)", asp = 1)
points(select(income, c("capital-gain", "capital-loss")), cex = (lof - 1) * 4, pch = 1, col = "red")
text(income[lof > 1.3,], labels = round(lof, 1)[lof > 1.3], pos = 3)

#education age
plot(select(income, c("education-num", "age")), pch = ".", main = "LOF (minPts = 10)", asp = 1)
points(select(income, c("education-num", "age")), cex = (lof - 1) * 4, pch = 1, col = "red")
text(income[lof > 1.3,], labels = round(lof, 1)[lof > 1.3], pos = 3)


```

##Outliers isolation forest
```{r}
library(isotree)
model = isolation.forest(income, ndim=1, ntrees=10)
scores = predict(model, income, type="score")
hist(scores, breaks = 10, main = "IF Scores")

plot(sort(scores), type = "l",  main = "IF Scores",
  xlab = "Points sorted by score", ylab = "IF score")

plot(select(income, c(`hours-per-week`, `education-num`)), pch = ".", main = "IF Scores", asp = 1)
points(select(income, c(`hours-per-week`, `education-num`))[scores > 0.5,], cex = as.data.frame(scores)[scores > 0.5,])
       
```
##Model Kmeans CLusters Kappa .57
```{r}
#income$income_above_50k <- as.factor(income$income_above_50k)
incomek$income_above_50k <- factor(incomek$income_above_50k, levels = c(FALSE, TRUE), labels = c("Below_50K", "Above_50K"))


#timer
start_time <- Sys.time()


# specify the model to be used (i.e. KNN, Naive Bayes, decision tree, random forest, bagged trees) and the tuning parameters used
ctrl <- trainControl(method = "cv", number = 3, classProbs = TRUE, summaryFunction = twoClassSummary)
set.seed(504) 

income_index <- createDataPartition(incomek$income_above_50k, p = 0.80, list = FALSE)
train <- incomek[ income_index, ]
test <- incomek[-income_index, ]

# example spec for rf
fit <- train(income_above_50k ~ .,
             data = train, 
             method = "rf",
             ntree = 20, 
             tuneLength = 3,
             metric = "ROC",
             trControl = ctrl)

fit

confusionMatrix(predict(fit, test),factor(test$income_above_50k))


#end timer
end_time <- Sys.time()
time_taken <- end_time - start_time
print(time_taken)
print(as.numeric(time_taken, units = "mins"))
```




##Model kappa .57 with no feature engineering
```{r}
#income$income_above_50k <- as.factor(income$income_above_50k)
income$income_above_50k <- factor(income$income_above_50k, levels = c(FALSE, TRUE), labels = c("Below_50K", "Above_50K"))


#timer
start_time <- Sys.time()


# specify the model to be used (i.e. KNN, Naive Bayes, decision tree, random forest, bagged trees) and the tuning parameters used
ctrl <- trainControl(method = "cv", number = 3, classProbs = TRUE, summaryFunction = twoClassSummary)
set.seed(504) 

income_index <- createDataPartition(income$income_above_50k, p = 0.80, list = FALSE)
train <- income[ income_index, ]
test <- income[-income_index, ]

# example spec for rf
fit <- train(income_above_50k ~ .,
             data = train, 
             method = "rf",
             ntree = 20, 
             tuneLength = 3,
             metric = "ROC",
             trControl = ctrl)

fit

confusionMatrix(predict(fit, test),factor(test$income_above_50k))


#end timer
end_time <- Sys.time()
time_taken <- end_time - start_time
print(time_taken)
print(as.numeric(time_taken, units = "mins"))
```


##EDA
```{r}
library(ggcorrplot)
library(ggplot2)
library(dplyr)

raw_income = read_csv("./openml_1590.csv", na=c("?"))

income2 = read_csv("./openml_1590.csv", na=c("?")) %>%
  drop_na() %>%
  mutate(income_above_50k = class==">50K") %>%
  select(-class)

##income2$income_above_50k <- as.numeric(factor(income2$income_above_50k, levels = unique(income2$income_above_50k))) - 1
income2$income_above_50k <- as.factor(income2$income_above_50k)
##try2
income2$income_above_50k <- as.numeric(income2$income_above_50k) - 1

#str(income2)

# Set the figure size for the plots
library(ggplot2)
options(repr.plot.width = 20, repr.plot.height = 12)

# Histograms for selected columns
hist_cols <- c("age", "fnlwgt", "education-num", "capital-gain", "capital-loss", "hours-per-week")
par(mfrow=c(2, 3))
for (col in hist_cols) {
  hist(income2[[col]], main=col, xlab=col, col="lightblue")
}

```


####Correlation Matrix Heatmap
```{r}

# Select the columns you want to include in the correlation matrix
selected_columns <- c("age", "fnlwgt", "education-num", "capital-gain", "capital-loss", "hours-per-week", "income_above_50k")

# Calculate the correlation matrix for the selected columns
corr_matrix <- cor(income2[, selected_columns])

# Create a heatmap of the correlation matrix with correlation coefficients
ggcorrplot(corr_matrix, lab = TRUE, lab_size = 4, tl.cex = 12, tl.col = "black", tl.srt = 45)


```

##Bucket Age
```{r}

# Bin the 'age' column
income2 <- income2 %>%
  mutate(age_group = case_when(
    age >= 0 & age <= 25 ~ "young",
    age > 25 & age <= 65 ~ "prime",
    age > 65 & age <= 100 ~ "retired"
  )) %>%
    select(-age)

# Reorder the age_group factor levels
income2$age_group <- factor(income2$age_group, levels = c("young", "prime", "retired"))

# Create the count plot with 'income_above_50k' as the hue
ggplot(income2, aes(x = age_group, fill = as.factor(income_above_50k))) +
  geom_bar(position = "dodge") +
  labs(x = "Age Group", y = "Count", fill = "Income Above 50K") +
  theme_minimal()

```

##Capital Diff
```{r}

#check if factor
##class(income2$income_above_50k)

# Convert 'income_above_50k' to a factor
##income2$income_above_50k <- as.factor(income2$income_above_50k)
##see above for same in EDA now

# Create 'Capital Diff' column and remove 'Capital Gain' and 'Capital Loss' columns
income2 <- income2 %>%
  mutate(capital_diff = `capital-gain` - `capital-loss`) %>%
  select(-`capital-gain`, -`capital-loss`)

# Bin the 'Capital Diff' column
income2 <- income2 %>%
  mutate(capital_diff = case_when(
    capital_diff >= -5000 & capital_diff <= 5000 ~ "Minor",
    capital_diff > 5000 & capital_diff <= 100000 ~ "Major"
  ))

# Create the count plot with 'Income' as the hue
ggplot(income2, aes(x = capital_diff, fill = income_above_50k)) +
  geom_bar(position = "dodge") +
  labs(x = "Capital Diff", y = "Count", fill = "Income") +
  theme_minimal()

```

##Drop Columns and without pay
```{r}
income2 <- income2 %>% select(-fnlwgt)

unique(income2$workclass)

income2 <- income2 %>%
  filter(workclass != "Without-pay")

```

##Bin Hours
```{r}
# Bin 'Hours per Week' column
income2 <- income2 %>%
  mutate(hours_per_week = case_when(
    `hours-per-week` >= 0 & `hours-per-week` <= 32 ~ "part_time",
    `hours-per-week` > 32 & `hours-per-week` <= 40 ~ "full_time",
    `hours-per-week` > 40 & `hours-per-week` <= 100 ~ "overtime"
  )) %>%
    select(-`hours-per-week`)

#Reorder hours per week
income2$hours_per_week <- factor(income2$hours_per_week, levels = c("part_time", "full_time", "overtime"))


# Create count plot with 'Income' as the hue
ggplot(income2, aes(x = hours_per_week, fill = income_above_50k)) +
  geom_bar(position = "dodge") +
  labs(x = "Hours per Week", y = "Count", fill = "Income") +
  theme_minimal()

```
##Looking at Education
```{r}

ggplot(raw_income, aes(x = education, fill = class)) +
  geom_bar(position = "dodge") +
  labs(x = "Education", y = "Count", fill = "Income") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
##Education same as education-num
```{r}
income2 %>%
  distinct(education, `education-num`) %>%
  arrange(`education-num`) %>%
  mutate(display = paste0("For ", education, ", the Education Number is ", `education-num`)) %>%
  pull(display) %>%
  print()

```
##Final Education counts
```{r}
income2 <- income2 %>%
  select(-`education-num`) %>%
  mutate(education = recode(education,
                            "11th" = "School",
                            "9th" = "School",
                            "7th-8th" = "School",
                            "5th-6th" = "School",
                            "10th" = "School",
                            "1st-4th" = "School",
                            "Preschool" = "School",
                            "12th" = "School"
                            ))

education_counts <- income2 %>%
  count(education) %>%
  arrange(desc(n))

education_counts

```
##Occupation
```{r}

# Calculate the count of 'income_above_50k == 1' for each occupation
occupation_order <- income2 %>%
  filter(income_above_50k == 1) %>%
  group_by(occupation) %>%
  summarize(count = n()) %>%
  arrange(-count) %>%
  pull(occupation)

# Reorder the 'occupation' factor levels based on the calculated order
income2$occupation_ordered <- factor(income2$occupation, levels = occupation_order)

# Create the bar plot
ggplot(income2, aes(x = occupation_ordered, fill = as.factor(income_above_50k))) +
  geom_bar(position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Occupation", y = "Count", fill = "Income Above 50K")

#remove occupation
income2 <- income2 %>%
    select(-occupation)
```
##Race
```{r}
ggplot(income2, aes(x = race, fill = as.factor(income_above_50k))) +
  geom_bar(position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  labs(x = "Race", y = "Count", fill = "Income Above 50K")


unique(income2$race)

income2 <- income2 %>%
  mutate(race = recode(race,
                       "Black" = "Other",
                       "Asian-Pac-Islander" = "Other",
                       "Amer-Indian-Eskimo" = "Other",
                       "Other" = "Other"
                       ))
```

##Sex vs Income
```{r}
ggplot(income2, aes(x = sex, fill = as.factor(income_above_50k))) +
  geom_bar(position = "dodge") +
  theme_minimal() +
  labs(x = "Sex", y = "Count", fill = "Income Above 50K")

```

##Country
```{r}
#Count of adults from each country
country_count <- table(income2$`native-country`)
barplot(country_count, xlab = "Countries", ylab = "Count", main = "Total adults from each Country", las = 2)

# Replace all countries except the first one with 'Other'
countries <- unique(income2$`native-country`)
income2$`native-country` <- ifelse(income2$`native-country` %in% countries[-1], "Other", income2$`native-country`)

#Country vs Income
ggplot(income2, aes(y = `native-country`, fill = as.factor(income_above_50k))) +
  geom_bar(position = "dodge") +
  theme_minimal() +
  labs(y = "Country", x = "Count", fill = "Income Above 50K")

```
##Create dummy variables
```{r}

income2_dummies <- income2 %>%
  dummy_cols(select_columns = setdiff(names(income2), "income_above_50k"), remove_selected_columns = T)

```




##EDA pt2 (education smaller bins)
```{r}

income3 <- income2 %>%
  mutate(education = recode(education,
                            "School" = "Some_college_orless",
                            "Some-college" = "Some_college_orless",
                            "HS-grad" = "Some_college_orless",
                            "Assoc-voc" = "Some_college_orless",
                            "Assoc-acdm" = "Some_college_orless",
                            "Masters" = "Master_plus",
                            "Prof-school" = "Master_plus",
                            "Doctorate" = "Master_plus"
                            ))

education_counts <- income3 %>%
  count(education) %>%
  arrange(desc(n))

education_counts

```

##Marital Status
```{r}
#unique names for marital status
unique(income3$`marital-status`)

#bin marital status
income3 <- income3 %>%
  mutate(`marital_status` = recode(`marital-status`,
                            "Never-married" = "never",
                            "Married-civ-spouse" = "married",
                            "Widowed" = "not_married",
                            "Separated" = "not_married",
                            "Divorced" = "not_married",
                            "Married-spouse-absent" = "married",
                            "Married-AF-spouse" = "married"
                            )) %>% 
    select(-`marital-status`)

```


##Tuesday kappa.57 RF full dataset
```{r eval=FALSE, message=FALSE, warning=FALSE}
# 
# income = read_csv("./openml_1590.csv", na=c("?")) %>%
#   drop_na() %>%
#   mutate(income_above_50k = class==">50K") %>%
#   select(-class) %>%
#   dummy_cols(remove_selected_columns = T)
# 
# #income$income_above_50k <- as.factor(income$income_above_50k)
# income$income_above_50k <- factor(income$income_above_50k, levels = c(FALSE, TRUE), labels = c("Below_50K", "Above_50K"))
# 
# 
# #timer
# start_time <- Sys.time()
# 
# 
# # specify the model to be used (i.e. KNN, Naive Bayes, decision tree, random forest, bagged trees) and the tuning parameters used
# ctrl <- trainControl(method = "cv", number = 3, classProbs = TRUE, summaryFunction = twoClassSummary)
# set.seed(504) 
# 
# income_index <- createDataPartition(income$income_above_50k, p = 0.80, list = FALSE)
# train <- income[ income_index, ]
# test <- income[-income_index, ]
# 
# #RF
# fit <- train(income_above_50k ~ .,
#              data = train, 
#              method = "rf",
#              ntree = 20, 
#              tuneLength = 3,
#              metric = "ROC",
#              trControl = ctrl)
# 
# fit
# 
# confusionMatrix(predict(fit, test),factor(test$income_above_50k))
# 
# 
# #end timer
# end_time <- Sys.time()
# time_taken <- end_time - start_time
# print(time_taken)
# print(as.numeric(time_taken, units = "mins"))
# 
# 
# # Naive Bayes
# fit_nb <- train(income_above_50k ~ .,
#                 data = train,
#                 method = "naive_bayes",
#                 metric = "ROC",
#                 trControl = ctrl)
# 
# fit_nb
# 
# confusionMatrix(predict(fit_nb, test), factor(test$income_above_50k))
# 
# #error
# # KNN
# #fit_knn <- train(income_above_50k ~ .,
# #                 data = train,
# #                 method = "knn",
# #                 tuneLength = 10,
# #                 metric = "ROC",
# #                 trControl = ctrl)
# 
# #fit_knn
# 
# #confusionMatrix(predict(fit_knn, test), factor(test$income_above_50k))
# 
# #error
# # K-Nearest Neighbors (KNN) with data normalization
# #fit_knn_normalized <- train(income_above_50k ~ .,
# #                            data = train,
# #                            method = "knn",
# #                            preProcess = "normalize", # Normalize the data
# #                            metric = "ROC",
# #                            trControl = ctrl)
# 
# #fit_knn_normalized
# 
# #confusionMatrix(predict(fit_knn_normalized, test), #factor(test$income_above_50k))
# 
# #error
# #library(rpart)
# # Decision Tree (CART)
# #fit_cart <- train(income_above_50k ~ .,
# #                  data = train,
# #                  method = "rpart",
# #                  metric = "ROC",
# #                  trControl = ctrl)
# 
# #fit_cart
# 
# #confusionMatrix(predict(fit_cart, test), factor(test$income_above_50k))
# 
# 
# # SVM Support Vector Machine
# fit_svm <- train(income_above_50k ~ .,
#                  data = train,
#                  method = "svmRadial",
#                  tuneLength = 3,
#                  metric = "ROC",
#                  trControl = ctrl)
# 
# fit_svm
# 
# confusionMatrix(predict(fit_svm, test), factor(test$income_above_50k))
# 
# # Logistic Regression Kappa .51
# fit_logreg <- train(income_above_50k ~ .,
#                     data = train,
#                     method = "glm",
#                     family = "binomial",
#                     metric = "ROC",
#                     trControl = ctrl)
# 
# fit_logreg
# 
# confusionMatrix(predict(fit_logreg, test), factor(test$income_above_50k))
# 
# # Gradient Boosting Machine Kappa .51
# fit_gbm <- train(income_above_50k ~ .,
#                  data = train,
#                  method = "gbm",
#                  tuneLength = 3,
#                  metric = "ROC",
#                  trControl = ctrl)
# 
# fit_gbm
# 
# confusionMatrix(predict(fit_gbm, test), factor(test$income_above_50k))
# 
# # XGBoost Kappa .53
# fit_xgb <- train(income_above_50k ~ .,
#                  data = train,
#                  method = "xgbTree",
#                  tuneLength = 3,
#                  metric = "ROC",
#                  trControl = ctrl)
# 
# fit_xgb
# 
# confusionMatrix(predict(fit_xgb, test), factor(test$income_above_50k))
# 

```

##Multiple Accuracy measurements (Skip)
```{r eval=FALSE, message=FALSE, warning=FALSE}

# ctrl <- trainControl(method = "cv", number = 5, classProbs = TRUE, summaryFunction = defaultSummary)
# 
# models <- list(
#   rf = train(income_above_50k ~ ., data = train, method = "rf", metric = "Accuracy", trControl = ctrl),
#   nb = train(income_above_50k ~ ., data = train, method = "nb", metric = "Accuracy", trControl = ctrl),
#   knn = train(income_above_50k ~ ., data = train, method = "knn", metric = "Accuracy", trControl = ctrl),
#   dt = train(income_above_50k ~ ., data = train, method = "rpart", metric = "Accuracy", trControl = ctrl)
# )
# 
# results <- resamples(models)
# summary(results, metric = "Accuracy")

```

##Multiple Kappa (Skip)
```{r eval=FALSE, message=FALSE, warning=FALSE}

# ctrl <- trainControl(method = "cv", number = 5, classProbs = TRUE, summaryFunction = defaultSummary)
# 
# models <- list(
#   rf = train(income_above_50k ~ ., data = train, method = "rf", metric = "Kappa", trControl = ctrl),
#   nb = train(income_above_50k ~ ., data = train, method = "nb", metric = "Kappa", trControl = ctrl),
#   knn = train(income_above_50k ~ ., data = train, method = "knn", metric = "Kappa", trControl = ctrl),
#   dt = train(income_above_50k ~ ., data = train, method = "rpart", metric = "Kappa", trControl = ctrl)
# )
# 
# results <- resamples(models)
# summary(results, metric = "Kappa")

```

##Multiple AUC (Skip)
```{r eval=FALSE, message=FALSE, warning=FALSE}

# ctrl <- trainControl(method = "cv", number = 5, classProbs = TRUE, summaryFunction = twoClassSummary)
# 
# # Train multiple models using different methods
# models <- list(
#   rf = train(income_above_50k ~ ., data = train, method = "rf", metric = "ROC", trControl = ctrl),
#   nb = train(income_above_50k ~ ., data = train, method = "nb", metric = "ROC", trControl = ctrl))#,
#   #knn = train(income_above_50k ~ ., data = train, method = "knn", metric = "ROC", trControl = ctrl),
#   #dt = train(income_above_50k ~ ., data = train, method = "rpart", metric = "ROC", trControl = ctrl)
# )
# 
# # Evaluate the models using the chosen metrics
# results <- resamples(models)
# 
# # Compare the models
# summary(results, metric = "ROC")
# 
# #error
# prob <- predict(fit, newdata=test)
# pred <- ifelse(prob > 0.5, 1, 0)
# confusionMatrix(factor(pred),factor(test$income_above_50k))
# #error
# library(pROC)
# myRoc <- roc(test$income_above_50k, prob)
# plot(myRoc)

```

##Multiple F1 (Skip)
```{r eval=FALSE, message=FALSE, warning=FALSE}
# 
# ctrl <- trainControl(method = "cv", number = 5, classProbs = TRUE, summaryFunction = prSummary)
# 
# models <- list(
#   rf = train(income_above_50k ~ ., data = train, method = "rf", metric = "F1", trControl = ctrl),
#   nb = train(income_above_50k ~ ., data = train, method = "nb", metric = "F1", trControl = ctrl),
#   knn = train(income_above_50k ~ ., data = train, method = "knn", metric = "F1", trControl = ctrl),
#   dt = train(income_above_50k ~ ., data = train, method = "rpart", metric = "F1", trControl = ctrl)
# )
# 
# results <- resamples(models)
# summary(results, metric = "F1")

```


##Tuesday pt2 EDA v1 variables kappa .41 RF
```{r eval=FALSE, message=FALSE, warning=FALSE}
# 
# income = read_csv("./openml_1590.csv", na=c("?")) %>%
#   drop_na() %>%
#   mutate(income_above_50k = class==">50K") %>%
#   select(-class) #%>%
#   #dummy_cols(remove_selected_columns = T)
# 
# #income$income_above_50k <- as.factor(income$income_above_50k)
# income$income_above_50k <- factor(income$income_above_50k, levels = c(FALSE, TRUE), labels = c("Below_50K", "Above_50K"))
# 
# #New adds
# 
# # Bin the 'age' column
# income <- income %>%
#   mutate(age_group = case_when(
#     age >= 0 & age <= 25 ~ "young",
#     age > 25 & age <= 65 ~ "prime",
#     age > 65 & age <= 100 ~ "retired"
#   )) %>%
#     select(-age)
# 
# # Create 'Capital Diff' column and remove 'Capital Gain' and 'Capital Loss' columns
# income <- income %>%
#   mutate(capital_diff = `capital-gain` - `capital-loss`) %>%
#   select(-`capital-gain`, -`capital-loss`)
# 
# # Bin the 'Capital Diff' column
# income <- income %>%
#   mutate(capital_diff = case_when(
#     capital_diff >= -5000 & capital_diff <= 5000 ~ "Minor",
#     capital_diff > 5000 & capital_diff <= 100000 ~ "Major"
#   ))
# 
# #drop Columns
# income <- income %>% select(-fnlwgt)
# 
# income <- income %>%
#   filter(workclass != "Without-pay")
# 
# # Bin 'Hours per Week' column
# income <- income %>%
#   mutate(hours_per_week = case_when(
#     `hours-per-week` >= 0 & `hours-per-week` <= 32 ~ "part_time",
#     `hours-per-week` > 32 & `hours-per-week` <= 40 ~ "full_time",
#     `hours-per-week` > 40 & `hours-per-week` <= 100 ~ "overtime"
#   )) %>%
#     select(-`hours-per-week`)
# 
# #Bin Education
# income <- income %>%
#   select(-`education-num`) %>%
#   mutate(education = recode(education,
#                             "11th" = "School",
#                             "9th" = "School",
#                             "7th-8th" = "School",
#                             "5th-6th" = "School",
#                             "10th" = "School",
#                             "1st-4th" = "School",
#                             "Preschool" = "School",
#                             "12th" = "School"
#                             ))
# 
# #Race
# income <- income %>%
#   mutate(race = recode(race,
#                        "Black" = "Other",
#                        "Asian-Pac-Islander" = "Other",
#                        "Amer-Indian-Eskimo" = "Other",
#                        "Other" = "Other"
#                        ))
# 
# # Replace all countries except the first one with 'Other'
# countries <- unique(income$`native-country`)
# income$`native-country` <- ifelse(income$`native-country` %in% countries[-1], "Other", income$`native-country`)
# 
# 
# #Create Dummy Variables
# income <- income %>%
#   dummy_cols(select_columns = setdiff(names(income), "income_above_50k"), remove_selected_columns = T)
# 
# #income %>%
#   #dummy_cols(remove_selected_columns = T)
# 
# #timer
# start_time <- Sys.time()
# 
# 
# # specify the model to be used (i.e. KNN, Naive Bayes, decision tree, random forest, bagged trees) and the tuning parameters used
# ctrl <- trainControl(method = "cv", number = 3, classProbs = TRUE, summaryFunction = twoClassSummary)
# set.seed(504) 
# 
# income_index <- createDataPartition(income$income_above_50k, p = 0.80, list = FALSE)
# train <- income[ income_index, ]
# test <- income[-income_index, ]
# 
# # example spec for rf
# fit <- train(income_above_50k ~ .,
#              data = train, 
#              method = "rf",
#              ntree = 20, 
#              tuneLength = 3,
#              metric = "ROC",
#              trControl = ctrl)
# 
# fit
# 
# confusionMatrix(predict(fit, test),factor(test$income_above_50k))
# 
# 
# #end timer
# end_time <- Sys.time()
# time_taken <- end_time - start_time
# print(time_taken)
# print(as.numeric(time_taken, units = "mins"))
```

##Tuesday pt3 w/EDA pt2 variables RF Kappa .46
```{r eval=FALSE, message=FALSE, warning=FALSE}
# 
# income = read_csv("./openml_1590.csv", na=c("?")) %>%
#   drop_na() %>%
#   mutate(income_above_50k = class==">50K") %>%
#   select(-class) #%>%
#   #dummy_cols(remove_selected_columns = T)
# 
# #income$income_above_50k <- as.factor(income$income_above_50k)
# income$income_above_50k <- factor(income$income_above_50k, levels = c(FALSE, TRUE), labels = c("Below_50K", "Above_50K"))
# 
# #New adds
# 
# # Bin the 'age' column
# income <- income %>%
#   mutate(age_group = case_when(
#     age >= 0 & age <= 25 ~ "young",
#     age > 25 & age <= 65 ~ "prime",
#     age > 65 & age <= 100 ~ "retired"
#   )) %>%
#     select(-age)
# 
# # Create 'Capital Diff' column and remove 'Capital Gain' and 'Capital Loss' columns
# income <- income %>%
#   mutate(capital_diff = `capital-gain` - `capital-loss`) %>%
#   select(-`capital-gain`, -`capital-loss`)
# 
# # Bin the 'Capital Diff' column
# income <- income %>%
#   mutate(capital_diff = case_when(
#     capital_diff >= -5000 & capital_diff <= 5000 ~ "Minor",
#     capital_diff > 5000 & capital_diff <= 100000 ~ "Major"
#   ))
# 
# #drop Columns
# income <- income %>% select(-fnlwgt)
# 
# income <- income %>%
#   filter(workclass != "Without-pay")
# 
# # Bin 'Hours per Week' column
# income <- income %>%
#   mutate(hours_per_week = case_when(
#     `hours-per-week` >= 0 & `hours-per-week` <= 32 ~ "part_time",
#     `hours-per-week` > 32 & `hours-per-week` <= 40 ~ "full_time",
#     `hours-per-week` > 40 & `hours-per-week` <= 100 ~ "overtime"
#   )) %>%
#     select(-`hours-per-week`)
# 
# #Bin Education
# income <- income %>%
#   select(-`education-num`) %>%
#   mutate(education = recode(education,
#                             "11th" = "School",
#                             "9th" = "School",
#                             "7th-8th" = "School",
#                             "5th-6th" = "School",
#                             "10th" = "School",
#                             "1st-4th" = "School",
#                             "Preschool" = "School",
#                             "12th" = "School"
#                             ))
# 
# #Race
# income <- income %>%
#   mutate(race = recode(race,
#                        "Black" = "Other",
#                        "Asian-Pac-Islander" = "Other",
#                        "Amer-Indian-Eskimo" = "Other",
#                        "Other" = "Other"
#                        ))
# 
# # Replace all countries except the first one with 'Other'
# countries <- unique(income$`native-country`)
# income$`native-country` <- ifelse(income$`native-country` %in% countries[-1], "Other", income$`native-country`)
# 
# 
# #Education further binned
# income <- income %>%
#   mutate(education = recode(education,
#                             "School" = "Some_college_orless",
#                             "Some-college" = "Some_college_orless",
#                             "HS-grad" = "Some_college_orless",
#                             "Assoc-voc" = "Some_college_orless",
#                             "Assoc-acdm" = "Some_college_orless",
#                             "Masters" = "Master_plus",
#                             "Prof-school" = "Master_plus",
#                             "Doctorate" = "Master_plus"
#                             ))
# 
# #occupation further binned
# income <- income %>%
#   mutate(occupation = recode(occupation,
#                             "Exec-managerial" = "high_tier",
#                             "Prof-specialty" = "high_tier",
#                             "Sales" = "mid_tier",
#                             "Craft-repair" = "mid_tier",
#                             "Adm-clerical" = "mid_tier",
#                             "Transport-moving" = "low_tier",
#                             "Tech-support" = "low_tier",
#                             "Machine-op-inspct" = "low_tier",
#                             "Protective-serv" = "low_tier",
#                             "Other-service" = "low_tier",
#                             "Farming-fishing" = "low_tier",
#                             "Handlers-cleaners" = "low_tier",
#                             "Armed-Forces" = "low_tier",
#                             "Priv-house-serv" = "low_tier"
#                             ))
# 
# #further binned marital status
# income <- income %>%
#   mutate(`marital_status` = recode(`marital-status`,
#                             "Never-married" = "never",
#                             "Married-civ-spouse" = "married",
#                             "Widowed" = "not_married",
#                             "Separated" = "not_married",
#                             "Divorced" = "not_married",
#                             "Married-spouse-absent" = "married",
#                             "Married-AF-spouse" = "married"
#                             )) %>% 
#     select(-`marital-status`)
# 
# 
# 
# 
# #Create Dummy Variables
# income <- income %>%
#   dummy_cols(select_columns = setdiff(names(income), "income_above_50k"), remove_selected_columns = T)
# 
# #income %>%
#   #dummy_cols(remove_selected_columns = T)
# 
# #timer
# start_time <- Sys.time()
# 
# 
# # specify the model to be used (i.e. KNN, Naive Bayes, decision tree, random forest, bagged trees) and the tuning parameters used
# ctrl <- trainControl(method = "cv", number = 3, classProbs = TRUE, summaryFunction = twoClassSummary)
# set.seed(504) 
# 
# income_index <- createDataPartition(income$income_above_50k, p = 0.80, list = FALSE)
# train <- income[ income_index, ]
# test <- income[-income_index, ]
# 
# # example spec for rf
# fit <- train(income_above_50k ~ .,
#              data = train, 
#              method = "rf",
#              ntree = 20, 
#              tuneLength = 3,
#              metric = "ROC",
#              trControl = ctrl)
# 
# fit
# 
# confusionMatrix(predict(fit, test),factor(test$income_above_50k))
# 
# 
# #end timer
# end_time <- Sys.time()
# time_taken <- end_time - start_time
# print(time_taken)
# print(as.numeric(time_taken, units = "mins"))
```


