---
title: "Machine Learning Project 3"
author: "Isaac Johnson & Jesse DeBolt"
date: "2023-04-10"
output: html_document
---

# EDA
### Setup
```{r}
library(tidyverse)
library(ggcorrplot)
library(ggplot2)
library(dplyr)
library(tidytext)
library(caret)
library(fastDummies)
library(randomForest)
library(broom)

raw_income = read_csv("openml_1590.csv", na=c("?"))

income2 = read_csv("openml_1590.csv", na=c("?")) %>%
  drop_na() %>%
  mutate(income_above_50k = class==">50K") %>%
  select(-class)

income2$income_above_50k <- as.numeric(factor(income2$income_above_50k, levels = unique(income2$income_above_50k))) - 1

str(income2)

```


### Bucket Age
```{r}

# Bin the 'age' column
income2 <- income2 %>%
  mutate(age_group = case_when(
    age >= 0 & age <= 25 ~ "young",
    age > 25 & age <= 65 ~ "prime",
    age > 65 & age <= 100 ~ "retired"
  )) %>%
    select(-age)

# Reorder the age_group factor levels
income2$age_group <- factor(income2$age_group, levels = c("young", "prime", "retired"))

```



### Bucket Capital gains/loss
```{r}

#check if factor
class(income2$income_above_50k)

# Convert 'income_above_50k' to a factor
income2$income_above_50k <- as.factor(income2$income_above_50k)

# Create 'Capital Diff' column and remove 'Capital Gain' and 'Capital Loss' columns
income2 <- income2 %>%
  mutate(capital_diff = `capital-gain` - `capital-loss`) %>%
  select(-`capital-gain`, -`capital-loss`)

# Bin the 'Capital Diff' column
income2 <- income2 %>%
  mutate(capital_diff = case_when(
    capital_diff >= -5000 & capital_diff <= 5000 ~ "Minor",
    capital_diff > 5000 & capital_diff <= 100000 ~ "Major"
  ))

```



### Drop Columns and without pay
```{r}
income2 <- income2 %>% select(-fnlwgt)

unique(income2$workclass)

income2 <- income2 %>%
  filter(workclass != "Without-pay")

```


### Bucket Hours
```{r}
# Bin 'Hours per Week' column
income2 <- income2 %>%
  mutate(hours_per_week = case_when(
    `hours-per-week` >= 0 & `hours-per-week` <= 32 ~ "part_time",
    `hours-per-week` > 32 & `hours-per-week` <= 40 ~ "full_time",
    `hours-per-week` > 40 & `hours-per-week` <= 100 ~ "overtime"
  )) %>%
    select(-`hours-per-week`)

#Reorder hours per week
income2$hours_per_week <- factor(income2$hours_per_week, levels = c("part_time", "full_time", "overtime"))

```


### Bucket Education
```{r}
income2 <- income2 %>%
  select(-`education-num`) %>%
  mutate(education = recode(education,
                            "11th" = "School",
                            "9th" = "School",
                            "7th-8th" = "School",
                            "5th-6th" = "School",
                            "10th" = "School",
                            "1st-4th" = "School",
                            "Preschool" = "School",
                            "12th" = "School"
                            ))

```


### Bucket Occupation
```{r}

# Calculate the count of 'income_above_50k == 1' for each occupation
occupation_order <- income2 %>%
  filter(income_above_50k == 1) %>%
  group_by(occupation) %>%
  summarize(count = n()) %>%
  arrange(-count) %>%
  pull(occupation)

# Reorder the 'occupation' factor levels based on the calculated order
income2$occupation_ordered <- factor(income2$occupation, levels = occupation_order)

```


### Bucket Race
```{r}

income2 <- income2 %>%
  mutate(race = recode(race,
                       "Black" = "Other",
                       "Asian-Pac-Islander" = "Other",
                       "Amer-Indian-Eskimo" = "Other",
                       "Other" = "Other"
                       ))
```


### Bucket Country
```{r}

# Replace all countries except the first one with 'Other'
countries <- unique(income2$`native-country`)
income2$`native-country` <- ifelse(income2$`native-country` %in% countries[-1], "Other", income2$`native-country`)

```


### Create dummy variables
```{r}

income2_dummies <- income2 %>%
  dummy_cols(select_columns = setdiff(names(income2), "income_above_50k"), remove_selected_columns = T)

```

## ----End EDA----

#PCA
###* Run PCA on the dataset.
```{r results='hide'}
pr_income <- prcomp(x = select(income2_dummies, -income_above_50k), scale = T, center = T)

summary(pr_income)

```

### Determine how many principal components are needed to explain half the variation
```{r}
cumulative_proportion <- cumsum(pr_income$sdev^2 / sum(pr_income$sdev^2))

n_components <- min(which(cumulative_proportion >= 0.5))

cat("The number of principal components needed to explain 50% of the variation is", n_components)

```
**Question:**Determine how many principal components are needed to explain half the variation
**Answer:** The number of principal components needed to explain 50% of the variation is 14


**Question:** Can you give some interpretation of the principal components?
### Viewing the loadings of each principal component. Loadings allow us to see the individual contribution of each predictor variable to each principal component.
```{r}
loadings <- rownames_to_column(as.data.frame(pr_income$rotation)) %>%
select(1:6) %>%
filter(abs(PC1) >= 0.25 | abs(PC2) >= 0.25 | abs(PC3) >= 0.25 | abs(PC4) >= 0.25 | abs(PC5) >= 0.25)

loadings

```
**Question:**Can you give some interpretation of the principal components?
**Answer:**pc1=Married-Male, pc2=Educated-Professional-Mexican, pc3=White-American-With_child, PC4=Mature-Single-Female, pc5=Black-Professional-American


### * Look at the scree plot. How many PCs would you choose based on that?
```{r}
screeplot(pr_income, type = "lines")

```
**Question:**Look at the scree plot. How many PCs would you choose based on that?
**Answer:** We would choose five PCs since it appears as though that is a good representation of the 'elbow'.


### Are the first few Principal Components good predictors of income_above_50k?
```{r}
# Add income above 50k back into the data and name principal components
prc <- bind_cols(select(income2_dummies,income_above_50k),as.data.frame(pr_income$x)) %>%
select(1:6) %>% 
  rename(Married_Male = PC1) %>% 
  rename(Educated_Professional_Mexican = PC2) %>% 
  rename(White_American_With_child = PC3) %>% 
  rename(Mature_Single_Female = PC4) %>% 
  rename(Black_Professional_American = PC5)

```

## Density Plots:

### Density of different components by income (Factors 1&2)
```{r}
prc %>%
  select(income_above_50k, Married_Male, Educated_Professional_Mexican) %>%
  pivot_longer(cols = -income_above_50k,names_to = "component",values_to = "loading") %>%
ggplot(aes(loading, fill=income_above_50k))+
geom_density(alpha=0.5)+
facet_grid(.~component)

```


### Density of different components by income (Factors 3&4)
```{r}
prc %>%
  select(income_above_50k, White_American_With_child, Mature_Single_Female) %>%
  pivot_longer(cols = -income_above_50k,names_to = "component",values_to = "loading") %>%
ggplot(aes(loading, fill=income_above_50k))+
geom_density(alpha=0.5)+
facet_grid(.~component)

```


### Density of different components by income (Factors 4&5)
```{r}
prc %>%
  select(income_above_50k, Mature_Single_Female, Black_Professional_American) %>%
  pivot_longer(cols = -income_above_50k,names_to = "component",values_to = "loading") %>%
ggplot(aes(loading, fill=income_above_50k))+
geom_density(alpha=0.5)+
facet_grid(.~component)

```

# Modeling
### Use these 5 factors for prediction
```{r}
fit <- train(income_above_50k ~ .,
             data = prc,
             method = "naive_bayes",
             trControl = trainControl(method = "cv"))


confusionMatrix(predict(fit, prc),factor(prc$income_above_50k))

```
**Question:**Are the first few Principal Components good predictors of income_above_50k?
**Answer:**With 82% accuracy we can say that we are somewhat accurate in predicting if income is above $50k. When reviewing the truth matrix:
Prediction     0     1
         0 30574  4757
         1  3421  6449
we can see that we are good at predicting true negatives, however predicting true positives are not as good.


**Question:** How well can you predict income_above_50k using the first 5 or 6 principal components? How about only the first 2?
```{r}
fit <- train(income_above_50k ~ Married_Male+Educated_Professional_Mexican,
             data = prc,
             method = "naive_bayes",
             trControl = trainControl(method = "cv"))


confusionMatrix(predict(fit, prc),factor(prc$income_above_50k))

```
**Question:** How well can you predict income_above_50k using the first 5 or 6 principal components? How about only the first 2?
**Answer:**For the first five or six principals we saw above that we are somewhat accurate in predicting if income is above \$50k. For the first two principal components we can see that we have an 82% accuracy, which means we are somewhat accurate in predicting if income is above \$50k.  When reviewing the truth matrix:
Prediction     0     1
         0 31106  5087
         1  2889  6119
we can see that we are slightly better at predicting true negatives, however predicting true positives are slightly better with the first five principal components.




